---
title: "Marine Aquaculture Suitability by EEZ Zone on West Coast of the U.S."
subtitle: "EDS 223 - Geospatial Analysis"
author: "Nicole Pepper"
date: last-modified
execute: 
  eval: true
  warning: false
  message: false
format:
  html:
    toc: true 
    code-fold: true
editor_options: 
  chunk_output_type: console

---
#### 1. Set up Workspace
```{r}
#| code-summary: "Load Libraries"

# ---- Load libraries ----
# Load libraries
library(terra)
library(geodata)
library(tidyverse)
library(tmap)
library(kableExtra)
library(spData)
library(spDataLarge)
library(here)
library(viridisLite) 
library(terra)
library(sf)
library(purrr)
library(stars)
library(raster)
library(dplyr)
library(testthat)
library(tigris)
```

```{r}
#| code-summary: "Read in Project Data"
#| output: false

# ---- Read in Project Data ----

# Create data folder object
data_folder <- here::here("data")

# Read in sea surface temperature (sst) data & make mean
sst_list <- list.files(data_folder,
                       pattern = "average_annual_sst_\\d{4}\\.tif$",
                       full.names = TRUE)
sst_stack <- lapply(sst_list, rast) %>%
  rast()

# Read in depth data
depth <- rast(here::here("data/depth.tif"))

# Read in EEZ Zones
eez_zones <- st_read(here::here("data/wc_regions_clean.shp"))


# Get US state boundaries from tigris library
states_sf <- tigris::states()

# Filter selected states
west_coast_states <- states_sf[states_sf$NAME %in% c("California", "Oregon", "Nevada", "Washington","Arizona","Idaho","Montana"), ] %>% st_transform(., st_crs(eez_zones))

```

#### 2. Prepare Data

```{r}
#| code-summary: "Prepare SST Data"

# Calculate mean SST
mean_sst <- mean(sst_stack)

# Convert SST from Kelvin to Celsius
mean_sst <- mean_sst - 273.15

```

```{r}
#| code-summary: "Prepare Data CRS"

# ---- Check CRS of data ----

# Create a warning if the data do not match
if (crs(depth) != crs(mean_sst)) {
  warning("The raster datasets do not have the same CRS.")
} else {
  message("The raster datasets are in the same CRS.")
}

# ---- Resample raster data to match resolution and extent ----

# Transform CRS
eez_zones <- st_transform(eez_zones, crs = crs(mean_sst)) 

# Resample depth data to match the resolution of SST with nearest neighbor
depth <- resample(depth, mean_sst, method = "near")

# Crop the resampled depth data to the extent of SST data
depth <- crop(depth, mean_sst, snap = "near")

# ---- Reproject depth to match mean_sst data ----

# Reproject depth layer
depth <- project(depth, crs(mean_sst))

# Create a warning if the data do not match
if (crs(depth) != crs(mean_sst)) {
  warning("The raster datasets do not have the same CRS.")
} else {
  message("The raster datasets are in the same CRS.")
}

# Stack data
data_stack <- c(depth, mean_sst)
names(data_stack) <- c("depth", "mean_sst")

```


#### 3. Calculate Oyster Habitat Suitability by EEZ Zone
```{r}
#| code-summary: "Reclassify SST & Depth Values for Suitable Habitat"

# ---- Assign suitable sst range ----
  # Define matrix for reclassification of sst
  sst <- matrix(c(-Inf, 11, 0,
                  11, 30, 1,
                  30, Inf, 0),
                ncol = 3, byrow = TRUE)
  sst_suitability <- terra::classify(data_stack[["mean_sst"]],
                                       rcl = sst,
                                       include.lowest = TRUE)

      
# ---- Assign suitable depth range ----

  # Define matrix for reclassification of depth
  depth <- matrix(c(-Inf, -70, 0,
                        -70, 0, 1,
                       0, Inf, 0),
                       ncol = 3, byrow = TRUE)
    
  depth_suitability <- terra::classify(data_stack[["depth"]],
                                         rcl = depth,
                                         include.lowest = TRUE)


```

```{r}
#| code-summary: "Calculate Suitability & Crop to EEZ Zone"

# ---- Calculate Suitability ----

  # Combine SST and depth suitability
  species_suitability <- sst_suitability * depth_suitability
    
  # Separate unsuitable area, reclassify 1 to NA
  species_unsuitable <- species_suitability
  species_unsuitable[species_unsuitable == 1] <- NA 
   
  # Remove unsuitable area, reclassify 0 to NA
  species_suitability[species_suitability == 0] <- NA
  
```

```{r}
#| code-summary: "Calculate Suitable Area By EEZ Zone"

  # Mask species_suitability to eez zones
  species_suitability <- mask(species_suitability, eez_zones)

  # Create a raster where each cell is it's size (will be used to summarize area)
  cell_area <- cellSize(species_suitability, unit = 'km')
  
  # Rasterize eez data by rgn field based on species_suitability resolution and extent
  eez_raster <- rasterize(eez_zones, species_suitability, field= 'rgn')
  
  # Mask the eez_raster to the species_suitability area
  eez_suitability <- mask(eez_raster, species_suitability)
    # Calculate the area
  eez_suitability_area <- zonal(cell_area, eez_suitability, fun = 'sum') %>%
    rename(suitable_area_km2 = area)
  
  # Join area calcs with the eez_zone dataset
  eez_suitability_area <- left_join(eez_zones, eez_suitability_area, by = 'rgn')  
  
  # Add column and sort by rank
  eez_suitability_area <- eez_suitability_area %>%
    mutate(rank = rank(-suitable_area_km2, ties.method = "min")) %>%
    arrange(rank)
```

```{r}
#| code-summary: "Visualize Preliminary Results"

# Set up the plotting layout: 1 row, 3 columns
par(mfrow = c(1, 3))

# Plot depth_suitability
plot(depth_suitability, main = "Depth Suitability")

# Plot sst_suitability
plot(sst_suitability, main = "SST Suitability")

# Plot eez_suitability
plot(eez_suitability, main = "EEZs Cropped to Suitable Areas")

# Reset plotting layout
par(mfrow = c(1, 1))

```

```{r}
#| code-summary: "Map Oyster Habitat Suitability by EEZ Zone"

# ---- Prepare labels for map ----

  # Add new line for spaces in eez region labels
  eez_suitability_area$rgn_label <- gsub(" ", "\n", eez_suitability_area$rgn)
  
  # Round area field to remove decimals and add scientific notation for area labels
  eez_suitability_area$suitable_area_km_rounded <- paste(format(round(eez_suitability_area$suitable_area_km), big.mark = ",", scientific = FALSE), "km²")

# ---- Create Suitability Map ----

  # Add unsuitable area for ocean background
  oyster_test_map <- tm_shape(species_unsuitable) + 
    tm_raster(palette = "#124D5C",
              legend.show = FALSE) +
  # Add eez_zones ranked by habitat suitability
  tm_shape(eez_suitability_area) +
    tm_fill(col = "rank",
            palette =  c("#00756D", "#00BFB2", "#6BCFC8", "#B2EFEF", "#CCFFFB"),
            title = "Suitable Area (km²)\n By EEZ Zone",
            border.alpha = 0.8,
            legend.show = FALSE) +
  # Add border color around EEZ zones boundaries 
  tm_shape(eez_suitability_area) +
    tm_borders(col = "#124D5C",
               lwd = .25) +
    
  # Add state borders for background
  tm_shape(west_coast_states) +
  tm_borders(col = "white",
             alpha = 0.1,
             lwd = .25) +
    
  tm_shape(species_suitability) +
  tm_raster(palette = "yellow",
              labels = "Suitable Area",
              title = '') + 
  
  # Add EEZ labels
  tm_shape(eez_suitability_area) +  
  tm_text("rgn_label",
            size = 0.6,
            col = "white",
            shadow = TRUE,
            fontface = "bold",
            ymod = .5) + 
    
  # Add suitable area labels
  tm_shape(eez_suitability_area) +
  tm_text("suitable_area_km_rounded",
            size = 0.7,
            col = "yellow",
            shadow = TRUE,
            fontface = "bold",
            ymod = -0.5) + 
  
  # Add tick marks 
  tm_grid(lines = FALSE, tick = TRUE, alpha = 0.8, n.x = 5, n.y = 5) +
  
  # Add legend for suitable area by EEZ Zone
  tm_add_legend(type = "fill", 
                col = c("#00756D", "#00BFB2", "#6BCFC8", "#B2EFEF", "#CCFFFB"), 
                labels = paste(eez_suitability_area$rank,eez_suitability_area$rgn_key), 
                title = "Exclusive Economic Zones (EEZ)\nRanked by Suitable Habitat Area",
                size = 1) +
  # Add compass
  tm_compass(type = "arrow",
             size = 1,
             position = c(0.1, 0.1),
             color.dark = "#07222B",
             color.light = "grey",
             text.color = "grey",
             text.size = .6) +
    
  # Format layout 
  tm_layout(main.title = paste("U.S. West Coast\n Oyster Habitat Suitability"),
              main.title.position = c("center", "top"),
              legend.text.color = "white",
              legend.position = c(0.55,0.45),
              legend.title.color = "white",
              legend.title.size = 1.2,
              bg.color = '#07222B')


```



### 4. Generalize Workflow for Habitat Suitability

#### Create a 'Choose-your-own species' function

**About the function:** 

Arguments:

- `species`: a string containing the species name

- `min_temp`: minimum suitable sea surface temperature in Celsius

- `min_temp`:  maximum suitable sea surface temperature in Celsius

- `min_depth`: minimum depth below sea level (0 = sea level)

- `max_depth`: maximum depth below sea level 


Outputs: Map of EEZ regions colored by amount of suitable area

```{r}
#| code-summary: "Create a Habitat Suitability Function"

# ---- Create suitability function ----
habitat_suitability_map <- function(species, min_sst, max_sst, min_depth, max_depth) {

# ---- Assign suitable sst range ----
  
  # Define matrix for reclassification of sst
  sst <- matrix(c(-Inf, min_sst, 0,
                  min_sst, max_sst, 1,
                  max_sst, Inf, 0),
                ncol = 3, byrow = TRUE)
  sst_suitability <- terra::classify(data_stack[["mean_sst"]],
                                       rcl = sst,
                                       include.lowest = TRUE)
# ---- Assign suitable depth range ----
  
  # Define matrix for reclassification of depth
  depth <- matrix(c(-Inf, max_depth, 0,
                        max_depth, min_depth, 1,
                       min_depth, Inf, 0),
                       ncol = 3, byrow = TRUE)
    
  depth_suitability <- terra::classify(data_stack[["depth"]],
                                      rcl = depth,
                                      include.lowest = TRUE)

# ---- Combine suitability ----

  # Combine SST and depth suitability
  species_suitability <- sst_suitability * depth_suitability
    
  # Separate unsuitable area, reclassify 1 to NA
  species_unsuitable <- species_suitability
  species_unsuitable[species_unsuitable == 1] <- NA 
   
  # Remove unsuitable area, reclassify 0 to NA
  species_suitability[species_suitability == 0] <- NA
   
  # Mask species_suitability to eez zones
  species_suitability <- mask(species_suitability, eez_zones)

# ---- Calculate Suitable Area By EEZ Zones ----

  # Create a raster where each cell is it's size (will be used to summarize area)
  cell_area <- cellSize(species_suitability, unit = 'km')
  
  # Rasterize eez data by rgn field based on species_suitability resolution and extent
  eez_raster <- rasterize(eez_zones, species_suitability, field= 'rgn')
  
  # Mask the eez_raster to the species_suitability area
  eez_suitability <- mask(eez_raster, species_suitability)

  # Calculate the area
  eez_suitability_area <- zonal(cell_area, eez_suitability, fun = 'sum') %>%
    rename(suitable_area_km2 = area)
  
  # Join area calcs with the eez_zone dataset
  eez_suitability_area <- left_join(eez_zones, eez_suitability_area, by = 'rgn')  
  
  # Add column and sort by rank
  eez_suitability_area <- eez_suitability_area %>%
    mutate(rank = rank(-suitable_area_km2, ties.method = "min")) %>%
    arrange(rank)

# ---- Prepare labels for map ----

  # Add new line for spaces in eez region labels
  eez_suitability_area$rgn_label <- gsub(" ", "\n", eez_suitability_area$rgn)
  
  # Round area field to remove decimals and add scientific notation for area labels
  eez_suitability_area$suitable_area_km_rounded <- paste(format(round(eez_suitability_area$suitable_area_km), big.mark = ",", scientific = FALSE), "km²")

# ---- Create Suitability Map ----

  # Add unsuitable area for ocean background
  tm_shape(species_unsuitable) + 
    tm_raster(palette = "#124D5C",
              legend.show = FALSE) +
  # Add eez_zones ranked by habitat suitability
  tm_shape(eez_suitability_area) +
    tm_fill(col = "rank",
            palette =  c("#00756D", "#00BFB2", "#6BCFC8", "#B2EFEF", "#CCFFFB"),
            title = "Suitable Area (km²)\n By EEZ Zone",
            border.alpha = 0.8,
            legend.show = FALSE) +
  # Add border color around EEZ zones boundaries 
  tm_shape(eez_suitability_area) +
    tm_borders(col = "#124D5C",
               lwd = .25) +
    
  # Add state borders for background
  tm_shape(west_coast_states) +
  tm_borders(col = "white",
             alpha = 0.1,
             lwd = .25) +
    
  tm_shape(species_suitability) +
  tm_raster(palette = "yellow",
              labels = "Suitable Area",
              title = '') + 
  
  # Add EEZ labels
  tm_shape(eez_suitability_area) +  
  tm_text("rgn_label",
            size = 0.6,
            col = "white",
            shadow = TRUE,
            fontface = "bold",
            ymod = .5) + 
    
  # Add suitable area labels
  tm_shape(eez_suitability_area) +
  tm_text("suitable_area_km_rounded",
            size = 0.7,
            col = "yellow",
            shadow = TRUE,
            fontface = "bold",
            ymod = -0.5) + 
  
  # Add tick marks 
  tm_grid(lines = FALSE, tick = TRUE, alpha = 0.8, n.x = 5, n.y = 5) +
  
  # Add legend for suitable area by EEZ Zone
  tm_add_legend(type = "fill", 
                col = c("#00756D", "#00BFB2", "#6BCFC8", "#B2EFEF", "#CCFFFB"), 
                labels = paste(eez_suitability_area$rank,eez_suitability_area$rgn_key), 
                title = "Exclusive Economic Zones (EEZ)\nRanked by Suitable Habitat Area",
                size = 1) +
  # Add compass
  tm_compass(type = "arrow",
             size = 1,
             position = c(0.1, 0.07),
             color.dark = "white",
             color.light = "grey",
             text.color = "grey",
             text.size = .6) +
    
  # Format layout 
  tm_layout(main.title = paste("U.S. West Coast\n",species, "Habitat Suitability"),
              main.title.position = c("center", "top"),
              legend.text.color = "white",
              legend.position = c(0.55,0.45),
              legend.title.color = "white",
              legend.title.size = 1.2,
              bg.color = '#07222B')
}

```

#### 5. Use function with Test Species


Table A.

| Species | Min Temp | Max Depth | Min Depth |  Max Depth |
|------------------|------------------|-------------------|------------------|------------------|
|Oyster          | 10                       | 20                        | - 40               |  0               | 
| Kelp          | 11                       | 30                         | -70                | 0               | 

Source: [sealifebase.ca](https://www.sealifebase.ca/search.php)
```{r}
#| code-summary: "Calculate Oyster Habitat Suitability"

habitat_suitability_map("Oyster", 11,30,0, -70)

```
*This map shows habitat suitability for oyster aquaculture along the U.S. West Coast. Suitable areas, highlighted in yellow, were calculated using thresholds for sea surface temperature and depth below sea level, based on their known range according to [sealifebase.ca](https://www.sealifebase.ca/search.php). The threshold values for oyster can be located above on Table A. The Exclusive Economic Zones (EEZ) regions are colored by the the amount of suitable area, where dark turquoise indicates the largest amount of area and light blue is the smallest. Central California ranked 1st on the list at 4,000 km² and Northern California ranked fifth, the lowest, at 178km².*

```{r}
#| code-summary: "Calculate Kelp Habitat Suitability"

habitat_suitability_map("Kelp", 10,20,0,-40)
```

*This map shows habitat suitability for giant kelp aquaculture along the U.S. West Coast. Suitable areas, highlighted in yellow, were calculated using thresholds for sea surface temperature and depth below sea level, based on their known range according to [sealifebase.ca](https://www.sealifebase.ca/search.php). The threshold values for kelp can be located above on Table A. The Exclusive Economic Zones (EEZ) regions are colored by the the amount of suitable area, where dark turquoise indicates the largest amount of area and light blue is the smallest. Central California ranked 1st on the list at 4,000 km² and Northern California ranked fifth, the lowest, at 178km².*


#### 6. Resources:

**Habitat Suitability Ranges:**
[sealifebase.ca](https://www.sealifebase.ca/search.php) Access date: 11/20/24

**Exclusive Economic Zones (EEZ) Boundaries:**
[Marine Regions](https://www.marineregions.org/eez.php) Access date: 11/20/24

**Sea Surface Temperature**
[NOAA’s 5km Daily Global Satellite Sea Surface Temperature Anomaly v3.1](https://coralreefwatch.noaa.gov/product/5km/index_5km_ssta.php) Access date: 11/20/24

**Bathymetry:**
[General Bathymetric Chart of the Oceans (GEBCO)](https://www.gebco.net/data_and_products/gridded_bathymetry_data/#area) Access date: 11/20/24